---
title: "US Birth Data"
author: "Sondre Løvås"
output: html_document
---

<!-- Data is loaded by the 'loadData' chunck. The code used to create the data has been set to not execute. 
To fully execute instead, set 'loadData' chunck eval and include to false and to true in data, extractData, 
sexProportion, averageWeight and sexProportionByDay chuncks.-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readr)
library(stringr)
library(lubridate)
library(zoo)
library(dplyr)
```

```{r data, eval=FALSE, include=FALSE}
d2017 = "data/Nat2017PublicUS.c20180516.r20180808.txt"
d2018 = "data/Nat2018PublicUS.c20190509.r20190717.txt"
d2019 = "data/Nat2019PublicUS.c20200506.r20200915.txt"
```

```{r loadData, eval=TRUE, include=TRUE}
load("data/df.RData")
load("data/dfSexProportion.RData")
load("data/dfAverageWeigth.RData")
load("data/dfSexProportionByDay.RData")
```

<!-- Extract year, month, day, sex and weight from the birth data. -->
<!-- Attributes location, year:9-12, month:13-14, day:23, sex:475, weight:504-507. -->
```{r extractData, eval=FALSE, include=FALSE}
extractData <- function(file) {
  data <- read_fwf(file,
                   fwf_positions(
                     c(9,13,23,475,504),
                     c(12,14,23,475,507),
                     c("Year", "Month", "Day", "Sex", "Weight")
                   )) %>%
          mutate(Date=make_date(Year, Month))
  return(data)
}

data2017 <- extractData(d2017)
data2018 <- extractData(d2018)
data2019 <- extractData(d2019)

df <- rbind(data2017, data2018, data2019)

```

<!-- Find the proportion of boys to girls. -->
```{r sexProportion, eval=FALSE, include=FALSE}
dfSexProportion <- as.data.frame(table(df$Date, df$Sex)) %>% 
  rename(., Date=Var1, Sex=Var2)
```

<!-- Find average weight. -->
```{r averageWeight, eval=FALSE, include=FALSE}
dfAverageWeight <- df
dfAverageWeight$Weight[dfAverageWeight$Weight == 9999] <- NA
dfAverageWeight$Weight <- as.numeric(as.character(df$Weight))
dfAverageWeight <- dfAverageWeight %>%
  group_by(Date, Sex) %>%
  summarise(AveWeigth=mean(Weight, na.rm = TRUE))
```

<!-- Find proportion of boys to girls by day of birth. -->
```{r sexProportionByDay, eval=FALSE, include=FALSE}
dfSexProportionByDay <- as.data.frame(table(df$Date, df$Sex, df$Day)) %>%
  rename(., Date=Var1, Sex=Var2, Day=Var3)
dfSexProportionByDay$Day <- recode(dfSexProportionByDay$Day,
                                   "1"="Sun", "2"="Mon",
                                   "3"="Tue", "4"="Wed",
                                   "5"="Thur", "6"="Fri",
                                   "7"="Sat")
```


In this report, US birth data such as frequency of boys to girls born and their average weights are visualized. 
The data is found at "https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm".
The data used are from the years 2017, 2018 and 2019.


<!-- Plot proportion of boys to girls -->
```{r plotSexProportion, message=FALSE, warning=FALSE, fig.width=15, fig.height=5}
ggplot(dfSexProportion,
       aes(x=month(Date), y=Freq, color=Sex, group=Sex)) + 
       geom_line() + 
       labs(x="Month", y="Frequency", color="Sex", title = "Frequency of boys and girls born by month in the years 2017, 2018 and 2019.") +
       scale_x_continuous(breaks = 1:12) +
       facet_wrap(~year(Date))
```

This plot shows how many boys and girls are born per month. Every year seems to have a big spike in August. 

<!-- Plot average weight -->
```{r plotAverageWeigth, message=FALSE, warning=FALSE, fig.width=15, fig.height=5}
ggplot(dfAverageWeight,
       aes(x=month(Date), y=AveWeigth, color=Sex, group=Sex)) + 
       geom_line() + 
       labs(x="Month", y="Weigth in grams", color="Sex", title = "Average birth weigth for boys and grils by month in the years 2017, 2018 and 2019.") +
       scale_x_continuous(breaks = 1:12) +
       facet_wrap(~year(Date))

```

Here it can be seen that boys weigh roughly 3350 grams and girls weigh 3225 grams on average when they are born. 


<!-- Plot proportion of boys to girls by day -->
```{r plotSexProportionByDay, message=FALSE, warning=FALSE, fig.width=16, fig.height=15}
ggplot(dfSexProportionByDay,
       aes(x=Day, y=Freq, color=Sex, group=Sex)) + 
       geom_line() + 
       labs(x="Day", y="Frequency", color="Sex", title = "Proportion of boys to girls by day in each month of the years 2017, 2018 and 2019") +
       facet_wrap(year(Date)~month(Date))
```
This plot shows how many boys and girls are born on average each day of every month in 2017, 2018 and 2019.




